Forcing for Isabelle

### 2018-12-28 LICS/intro_m.tex
Intro de Miguel

After G\"odel's Incompleteness Theorems, we cannot expect to have a
formal proof of the consistency of Set Theory in $ZF$. Besides its own
consistency, there are other results which are undecided by $ZF$: the
undecidability of Continuum Hypothesis lead to the development of
techniques for independence proofs. First G\"odel introduced the
theory of \emph{inner models}, which gives rise to his model $L$ of
the \emph{Axiom of Constructibility} \cite{godel-L} and proved the
relative consistency of the Axiom of Choice and the Generalized
Continuum Hypothesis with $\ZF$. Thirty years later Paul
J. Cohen~\cite{Cohen-CH-PNAS} devised the technique of \emph{forcing},
which is the only known way of \emph{extending} models of $\ZF$; in
particular, it can be used to prove the relative consistency of the
negation of the Continuum Hypothesis.

In this work we address a substantial part of formalizing the proof
that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
given by forcing also satisfies $\ZF$. As remarked by
\citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
  is a model for set theory, the hardest axiom to verify is
  Comprehension.}  The most important achievement of this paper is the
mechanization in Isabelle of this result; en route to this, we also
formalized the satisfaction by $M[G]$ of Extensionality, Foundation,
Union, Infinity.

Our development benefited from the extraordinary work done by Lawrence
Paulson \cite{paulson_2003} on the formalization of G\"odel's
constructible universe in the proof assistant \emph{Isabelle}. In the
spirit of the \emph{formal abstract} project~\cite{hales-fabstracts},
we explicit the technical assumptions of forcing that are involved in
the proof of Separation for $M[G]$. It seems that this work is
\emph{Es un lindo test a la capacidad de implementación formal de
  matemática con la tecnología actual}.

\fbox{Acá viene una oración introduciendo lo que sigue.}

@@

\fbox{El párrafo que sigue lo sacaría}

In this work we address a substantial part the problem of formalizing
the proof that given a model of $\ZF$, its extensions by forcing, 
\emph{generic extensions},
satisfy the same axioms. In particular, the first axiom that
genuinely calls for a use of the forcing machinery is that of
\emph{Separation} (\emph{Aussonderungsaxiom}), otherwise known as
``Axiom of (Restricted) Comprehension''  or ``Specification.'' In
\cite{2018arXiv180705174G} it was shown that  generic extensions of
models of $\ZF$ 
satisfy the Axiom of Pairing, and in this work we show that also
Extensionality, Foundation, Union, Infinity (under extra assumptions),
and Separation go through. The
Axiom of Powerset is not treated in the present paper but it poses no
extra difficulties; we discuss the Axiom of Replacement in the
conclusions.

The proof that these axioms hold in generic extension is independent
of the \emph{proofs} of the ``fundamental'' theorems of forcing. In
particular, one of the most complex parts of those proofs is the
definition of the \emph{forcing relation}. It turns out that the
particular (``implementation'') details of this  definition  does not
appear in the statement of the fundamental theorems, and our
formalization works inside a \emph{locale} (set of assumptions) that
includes the statement of these theorems. The formalization of the
fundamental theorems of forcing roughly comprises little less than a
half of our full project.

>>>>

After G\"odel's Incompleteness Theorems, we cannot expect to have a
formal proof of the consistency of Set Theory in $ZF$. Besides its own
consistency, there are other results which are undecided by $ZF$: the
undecidability of Continuum Hypothesis leads to the development of
techniques for independence proofs. First G\"odel introduced the
theory of \emph{inner models}, which gives rise to his model $L$ of
the \emph{Axiom of Constructibility} \cite{godel-L} and proved the
relative consistency of the Axiom of Choice and the Generalized
Continuum Hypothesis with $\ZF$. Thirty years later Paul
J. Cohen~\cite{Cohen-CH-PNAS} devised the technique of \emph{forcing},
which is the only known way of \emph{extending} models of $\ZF$; in
particular, it can be used to prove the relative consistency of the
negation of the Continuum Hypothesis.

In this work we address a substantial part of formalizing the proof
that given a model $M$ of $\ZF$, any \emph{generic extension} $M[G]$
given by forcing also satisfies $\ZF$. As remarked by
\citet[][p.250]{kunen2011set} \enquote{[...] in verifying that $M[G]$
  is a model for set theory, the hardest axiom to verify is
  Comprehension.}  The most important achievement of this paper is the
mechanization in Isabelle of a substantial part of this result; en
route to this, we also formalized the satisfaction by $M[G]$ of
Extensionality, Foundation, Union, and Infinity.

Our development benefited from the extraordinary work done by Lawrence
Paulson \cite{paulson_2003} on the formalization of G\"odel's
constructible universe in the proof assistant \emph{Isabelle}. The
ultimate goal of our project is the formalization of forcing to
complete the mechanization of the independence of the Continuum
Hypothesis. We think that this project constitutes an interesting case
which stresses how feasible is to formally implement mathematics that
involve several leves of reasoning.

The \emph{Formal Abstract} project~\cite{hales-fabstracts} proposes
that the mechanization of mathematics can advance by formalizing some
main result under the explicit assumption of the material
(definitions, propositions, lemmas) on which it is based. In line with
this advice, we profited that the proof that those axioms hold in
generic extension is independent of the \emph{proofs} of the
``fundamental'' theorems of forcing. Let us remark that the definition
of the forces relation is, by itself, quite demanding; the
formalization of it and of the fundamental theorems of forcing roughly
comprises little less than a half of our full project.

\fbox{Parece mucho comienzo sólo para introducir a Kunen}
\fbox{¿lo puedo achurar un poco?}

@@
